---
title: "Fungal Treatment Analysis Overview"
format: 
  html:
    toc: false
    number-sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      fig.width = 12, fig.height = 8, dpi = 300)

library(tidyverse)
library(here)
library(ggplot2)
library(scales)
library(knitr)
library(patchwork)

# Load data
exp_data <- read_csv(here("data/processed/experiment_final.csv"), show_col_types = FALSE) %>%
  mutate(
    species = ifelse(species == "ctrl", "Control", species),
    MAX_PLATE_AREA = 63.62
  )

# Remove F37 Mold and F39 Thermo tolerant from analysis
exp_data <- exp_data %>%
  filter(species != "F37 Mold", species != "F39 Thermo tolerant")

# Remove data points marked with "exclude" during data cleaning
exp_data <- exp_data %>%
  filter(is.na(exclude) | exclude != "exclude")

# Add growth status classification and biological replicate identifier
exp_data <- exp_data %>%
  mutate(
    growth_status = case_when(
      species == "Control" ~ "Control",
      is.na(area_size_14dpi) | area_size_14dpi < 1 ~ "No Growth",
      TRUE ~ "Growth Observed"
    ),
    biological_replicate = paste(id_faeces, starting_date, sep = "_")
  )

# Create consistent species ordering with Control first
species_order <- exp_data %>%
  distinct(species) %>%
  arrange(species != "Control", species) %>%
  pull(species)

exp_data <- exp_data %>%
  mutate(species = factor(species, levels = species_order))

# Function to aggregate technical replicates
aggregate_technical_reps <- function(data) {
  data %>%
    group_by(species, experiment, biological_replicate, growth_status) %>%
    summarise(
      # Growth data
      area_size_7dpi_mean = mean(area_size_7dpi, na.rm = TRUE),
      area_size_14dpi_mean = mean(area_size_14dpi, na.rm = TRUE),
      
      # pH data
      ph_0dpi_mean = mean(ph_0dpi, na.rm = TRUE),
      ph_14dpi_mean = mean(ph_14dpi, na.rm = TRUE),
      ph_change_mean = mean(ph_14dpi - ph_0dpi, na.rm = TRUE),
      
      # Weight data
      weight_0dpi_mean = mean(weight_0dpi, na.rm = TRUE),
      weight_14dpi_mean = mean(weight_14dpi, na.rm = TRUE),
      weight_percent_change_mean = mean(((weight_14dpi - weight_0dpi) / weight_0dpi) * 100, na.rm = TRUE),
      
      # E. coli data
      ecoli_conc_mean_start = mean(ecoli_conc_mean, na.rm = TRUE),
      ecoli_conc_14dpi_mean = mean(ecoli_conc_14dpi, na.rm = TRUE),
      ecoli_log_reduction_mean = mean(
        ifelse(ecoli_conc_mean > 0 & ecoli_conc_14dpi > 0,
               log10(ecoli_conc_mean) - log10(ecoli_conc_14dpi),
               NA),
        na.rm = TRUE
      ),
      
      n_technical = n(),
      .groups = "drop"
    )
}

# Create aggregated dataset
exp_data_avg <- aggregate_technical_reps(exp_data)
```

# Sample Size Summary

## Experiment 1_1 Replicates

```{r exp1-replicates}
exp1_summary <- exp_data %>%
  filter(experiment == "1_1") %>%
  group_by(species) %>%
  summarise(
    technical_replicates = n(),
    biological_replicates = n_distinct(paste(id_faeces, starting_date)),
    growth_data = sum(!is.na(area_size_14dpi)),
    ecoli_data = sum(!is.na(ecoli_conc_14dpi)),
    weight_data = sum(!is.na(weight_14dpi)),
    .groups = "drop"
  ) %>%
  arrange(desc(technical_replicates))

kable(exp1_summary,
      caption = "**Experiment 1_1: Sample sizes and data availability**",
      col.names = c("Species", "Technical Reps", "Biological Reps", "Growth Data", "E.coli Data", "Weight Data"))
```

## Experiment 1_2 Replicates

```{r exp2-replicates}
exp2_summary <- exp_data %>%
  filter(experiment == "1_2") %>%
  group_by(species) %>%
  summarise(
    technical_replicates = n(),
    biological_replicates = n_distinct(paste(id_faeces, starting_date)),
    growth_data = sum(!is.na(area_size_14dpi)),
    ecoli_data = sum(!is.na(ecoli_conc_14dpi)),
    weight_data = sum(!is.na(weight_14dpi)),
    .groups = "drop"
  ) %>%
  arrange(desc(technical_replicates))

kable(exp2_summary,
      caption = "**Experiment 1_2: Sample sizes and data availability**",
      col.names = c("Species", "Technical Reps", "Biological Reps", "Growth Data", "E.coli Data", "Weight Data"))
```

# Growth Success Rate Analysis

## Growth Success by Species

```{r growth-success-rate, fig.height=8, fig.width=10}
# Calculate growth success rates
growth_success <- exp_data %>%
  filter(species != "Control") %>%  # Exclude control as it's not expected to grow
  group_by(species, experiment) %>%
  summarise(
    total_attempts = n(),
    successful_growth = sum(growth_status == "Growth Observed"),
    success_rate = (successful_growth / total_attempts) * 100,
    .groups = "drop"
  )

# Create visualization
ggplot(growth_success, aes(x = species, y = success_rate, fill = experiment)) +
  geom_bar(stat = "identity", position = position_dodge(0.8), alpha = 0.8) +
  geom_text(aes(label = paste0(round(success_rate, 0), "%\n(", successful_growth, "/", total_attempts, ")")),
            position = position_dodge(0.8), vjust = -0.5, size = 3) +
  scale_y_continuous(limits = c(0, 110), breaks = seq(0, 100, 20)) +
  labs(title = "Fungal Growth Success Rate by Species",
       x = "Species",
       y = "Growth Success Rate (%)",
       fill = "Experiment") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.minor = element_blank())
```

**Key Observations:** - Some species show consistent growth across experiments while others are more variable - Growth success rates help identify reliable strains for treatment applications - Failed growth attempts should be considered when interpreting pH, weight, and E. coli data

# Growth Analysis

## Experiment 1_1 Growth

```{r growth-exp1, fig.height=8}
# Growth data for Experiment 1_1 - raw data for boxplots
growth_exp1_raw <- exp_data %>%
  filter(experiment == "1_1") %>%
  select(species, area_size_7dpi, area_size_14dpi) %>%
  pivot_longer(cols = c(area_size_7dpi, area_size_14dpi),
               names_to = "timepoint",
               values_to = "area_cm2") %>%
  filter(!is.na(area_cm2), is.finite(area_cm2)) %>%
  mutate(
    day = case_when(
      timepoint == "area_size_7dpi" ~ "7 DPI",
      timepoint == "area_size_14dpi" ~ "14 DPI"
    ),
    species = factor(species, levels = species_order)
  )

if(nrow(growth_exp1_raw) > 0) {
  # Add separation lines between species
  n_species <- length(levels(growth_exp1_raw$species))
  separation_lines <- seq(1.5, n_species - 0.5, by = 1)
  
  ggplot(growth_exp1_raw, aes(x = area_cm2, y = species, fill = day)) +
    geom_hline(yintercept = separation_lines, color = "gray80", linetype = "solid", size = 0.5) +
    geom_boxplot(alpha = 0.7, position = position_dodge(width = 0.8)) +
    geom_vline(xintercept = 63.62, linetype = "dashed", color = "red", alpha = 0.7) +
    labs(title = "Experiment 1_1: Fungal Growth at 7 and 14 DPI",
         x = "Colony Area (cm²)",
         y = "Species",
         fill = "Timepoint") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank())
} else {
  plot.new()
  text(0.5, 0.5, "No growth data available for Experiment 1_1", cex = 1.2)
}
```

## Experiment 1_2 Growth

```{r growth-exp2, fig.height=8}
# Growth data for Experiment 1_2 - raw data for boxplots
growth_exp2_raw <- exp_data %>%
  filter(experiment == "1_2") %>%
  select(species, area_size_7dpi, area_size_14dpi) %>%
  pivot_longer(cols = c(area_size_7dpi, area_size_14dpi),
               names_to = "timepoint",
               values_to = "area_cm2") %>%
  filter(!is.na(area_cm2), is.finite(area_cm2)) %>%
  mutate(
    day = case_when(
      timepoint == "area_size_7dpi" ~ "7 DPI",
      timepoint == "area_size_14dpi" ~ "14 DPI"
    ),
    species = factor(species, levels = species_order)
  )

if(nrow(growth_exp2_raw) > 0) {
  # Add separation lines between species
  n_species <- length(levels(growth_exp2_raw$species))
  separation_lines <- seq(1.5, n_species - 0.5, by = 1)
  
  ggplot(growth_exp2_raw, aes(x = area_cm2, y = species, fill = day)) +
    geom_hline(yintercept = separation_lines, color = "gray80", linetype = "solid", size = 0.5) +
    geom_boxplot(alpha = 0.7, position = position_dodge(width = 0.8)) +
    geom_vline(xintercept = 63.62, linetype = "dashed", color = "red", alpha = 0.7) +
    labs(title = "Experiment 1_2: Fungal Growth at 7 and 14 DPI",
         x = "Colony Area (cm²)",
         y = "Species",
         fill = "Timepoint") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank())
} else {
  plot.new()
  text(0.5, 0.5, "No growth data available for Experiment 1_2", cex = 1.2)
}
```

Several fungi reliably grow on fresh human feces, with Sordaria isolates being the fastest-growing species on both pure feces and a 2:1 mixture with wood chips. While the fungal mycelium in unmixed samples degrades in about seven days, it remains visible in the mixed samples for at least 14 days, only becoming weaker over time. The formation of conidia (Mucor, Trichoderma) and fruiting bodies (perithecia, Sordaria) is observed exclusively in the mixed samples. One Basidiomycete isolate (F40, possibly Coprinopsis) showed strong mycelial growth in one assay but failed in another. To improve the consistency of this isolate's growth, adding more accessible cellulose and using a stronger inoculum could be beneficial.

# pH Changes

## Experiment 1_1 pH Changes

```{r ph-exp1, fig.height=6, fig.width=10}
ph_exp1 <- exp_data_avg %>%
  filter(experiment == "1_1", !is.na(ph_0dpi_mean), !is.na(ph_14dpi_mean))

if(nrow(ph_exp1) > 0) {
  # Only show growth status if there's variation - otherwise just use species
  ph_exp1_clean <- ph_exp1 %>%
    filter(growth_status %in% c("Control", "Growth Observed")) %>%  # Remove "No Growth" if no variance
    mutate(display_group = case_when(
      species == "Control" ~ "Control",
      growth_status == "Growth Observed" ~ paste(species, "(Growth)")
    ))
  
  ggplot(ph_exp1_clean, aes(x = reorder(display_group, ph_change_mean, median), y = ph_change_mean)) +
    geom_boxplot(aes(fill = ifelse(species == "Control", "Control", "Fungal Treatment")), 
                 alpha = 0.7, outlier.size = 1) +
    geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
    scale_fill_manual(values = c("Control" = "gray60", "Fungal Treatment" = "lightblue")) +
    labs(title = "pH Changes from 0 to 14 DPI (Experiment 1_1)",
         x = "",
         y = "pH Change",
         fill = "") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
          panel.grid.minor = element_blank(),
          legend.position = "top")
} else {
  plot.new()
  text(0.5, 0.5, "No pH data available", cex = 1.2)
}
```

**Key Finding:** Successful fungal growth consistently increases pH by 3-4 units compared to controls.

## Experiment 1_2 pH Changes

```{r ph-exp2, fig.height=8}
ph_exp2 <- exp_data_avg %>%
  filter(experiment == "1_2", !is.na(ph_0dpi_mean), !is.na(ph_14dpi_mean))

if(nrow(ph_exp2) > 0) {
  # Create plot with growth status
  p2 <- ggplot(ph_exp2, aes(x = factor(species, levels = rev(species_order)), y = ph_change_mean)) +
    geom_boxplot(aes(fill = growth_status), alpha = 0.7, position = position_dodge(0.8)) +
    geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
    scale_fill_manual(values = c("Control" = "gray50", "Growth Observed" = "lightblue", "No Growth" = "lightcoral")) +
    coord_flip() +
    labs(title = "Experiment 1_2: pH Changes from 0 to 14 DPI (Biological Replicates)",
         subtitle = "Averaged across technical replicates, split by growth status",
         x = "Species",
         y = "pH Change (14 DPI - 0 DPI)",
         fill = "Growth Status") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          legend.position = "bottom")
  
  # Summary statistics by growth status
  summary_stats <- ph_exp2 %>%
    group_by(growth_status) %>%
    summarise(
      n = n(),
      mean_ph_change = mean(ph_change_mean, na.rm = TRUE),
      sd_ph_change = sd(ph_change_mean, na.rm = TRUE),
      median_ph_change = median(ph_change_mean, na.rm = TRUE)
    )
  
  print(p2)
  
  # Print summary table
  kable(summary_stats,
        caption = "**pH Change Summary by Growth Status**",
        col.names = c("Growth Status", "N", "Mean pH Change", "SD", "Median"),
        digits = 2)
  
} else {
  plot.new()
  text(0.5, 0.5, "No pH data available for Experiment 1_2", cex = 1.2)
}
```

-   In pure and mixed samples the pH of all plates that showed fungal growth increased to around 9
-   Plates without fungal growth the pH was lower around 5.5

This change of pH is due to ammonia production, ammonia is produced due to bacterial metabolism of amino acids to gain energy. The problem is that fungal enzymes might not work at this pH and the degradation is slowed down.

Fungi are known to produce acids and during degradation and prefer an environment slightly acidic around 5.5

Follow up question Can the pH be kept in the optimal range by replacing the hard degradable wood chips with easier degradable material flax fibers and flax fibers mixed with cellulose powder.

# E. coli Reduction

## Experiment 1_1 E. coli Reduction

```{r ecoli-exp1, fig.height=8}
ecoli_exp1 <- exp_data_avg %>%
  filter(experiment == "1_1", !is.na(ecoli_conc_mean_start), !is.na(ecoli_conc_14dpi_mean),
         ecoli_conc_mean_start > 0, ecoli_conc_14dpi_mean > 0)

if(nrow(ecoli_exp1) > 0) {
  # Create plot with growth status
  p3 <- ggplot(ecoli_exp1, aes(x = factor(species, levels = rev(species_order)), y = ecoli_log_reduction_mean)) +
    geom_boxplot(aes(fill = growth_status), alpha = 0.7, position = position_dodge(0.8)) +
    geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
    scale_fill_manual(values = c("Control" = "gray50", "Growth Observed" = "lightgreen", "No Growth" = "lightcoral")) +
    coord_flip() +
    labs(title = "Experiment 1_1: E. coli Log Reduction (Biological Replicates)",
         subtitle = "Averaged across technical replicates, split by growth status",
         x = "Species",
         y = "Log Reduction (Log10 Initial - Log10 Final)",
         fill = "Growth Status") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          legend.position = "bottom")
  
  # Summary statistics by growth status
  summary_stats <- ecoli_exp1 %>%
    group_by(growth_status) %>%
    summarise(
      n = n(),
      mean_log_reduction = mean(ecoli_log_reduction_mean, na.rm = TRUE),
      sd_log_reduction = sd(ecoli_log_reduction_mean, na.rm = TRUE),
      median_log_reduction = median(ecoli_log_reduction_mean, na.rm = TRUE)
    )
  
  print(p3)
  
  # Print summary table
  kable(summary_stats,
        caption = "**E. coli Log Reduction Summary by Growth Status**",
        col.names = c("Growth Status", "N", "Mean Log Reduction", "SD", "Median"),
        digits = 2)
  
} else {
  plot.new()
  text(0.5, 0.5, "No E. coli data available for Experiment 1_1", cex = 1.2)
}
```

## Experiment 1_2 E. coli Reduction

```{r ecoli-exp2, fig.height=8}
ecoli_exp2 <- exp_data_avg %>%
  filter(experiment == "1_2", !is.na(ecoli_conc_mean_start), !is.na(ecoli_conc_14dpi_mean),
         ecoli_conc_mean_start > 0, ecoli_conc_14dpi_mean > 0)

if(nrow(ecoli_exp2) > 0) {
  # Create plot with growth status
  p4 <- ggplot(ecoli_exp2, aes(x = factor(species, levels = rev(species_order)), y = ecoli_log_reduction_mean)) +
    geom_boxplot(aes(fill = growth_status), alpha = 0.7, position = position_dodge(0.8)) +
    geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
    scale_fill_manual(values = c("Control" = "gray50", "Growth Observed" = "lightgreen", "No Growth" = "lightcoral")) +
    coord_flip() +
    labs(title = "Experiment 1_2: E. coli Log Reduction (Biological Replicates)",
         subtitle = "Averaged across technical replicates, split by growth status",
         x = "Species",
         y = "Log Reduction (Log10 Initial - Log10 Final)",
         fill = "Growth Status") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          legend.position = "bottom")
  
  # Summary statistics by growth status
  summary_stats <- ecoli_exp2 %>%
    group_by(growth_status) %>%
    summarise(
      n = n(),
      mean_log_reduction = mean(ecoli_log_reduction_mean, na.rm = TRUE),
      sd_log_reduction = sd(ecoli_log_reduction_mean, na.rm = TRUE),
      median_log_reduction = median(ecoli_log_reduction_mean, na.rm = TRUE)
    )
  
  print(p4)
  
  # Print summary table
  kable(summary_stats,
        caption = "**E. coli Log Reduction Summary by Growth Status**",
        col.names = c("Growth Status", "N", "Mean Log Reduction", "SD", "Median"),
        digits = 2)
  
} else {
  plot.new()
  text(0.5, 0.5, "No E. coli data available for Experiment 1_2", cex = 1.2)
}
```

When comparing samples with and without fungal growth, the presence of fungi led to an increase in E. coli numbers in both pure and mixed fecal samples. This is likely because the fungi's enzymatic activity makes nutrients like proteins, lipids, and lignocellulose more accessible to the bacteria. This accelerated degradation process boosts the overall microbial activity of the existing bacteria. The utilization of carbon by bacteria also causes a rise in pH due to the degradation of amino acids. Overall, the presence of fungi appears to increase the general bacterial richness compared to samples without fungal growth.

# Weight Changes

## Experiment 1_1 Weight Changes

### Wet Weight Changes

```{r wet-weight-exp1, fig.height=8}
wet_weight_exp1 <- exp_data_avg %>%
  filter(experiment == "1_1", !is.na(weight_0dpi_mean), !is.na(weight_14dpi_mean))

if(nrow(wet_weight_exp1) > 0) {
  # Create plot with growth status
  p5 <- ggplot(wet_weight_exp1, aes(x = factor(species, levels = rev(species_order)), y = weight_percent_change_mean)) +
    geom_boxplot(aes(fill = growth_status), alpha = 0.7, position = position_dodge(0.8)) +
    geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
    scale_fill_manual(values = c("Control" = "gray50", "Growth Observed" = "lightsteelblue", "No Growth" = "lightcoral")) +
    coord_flip() +
    labs(title = "Experiment 1_1: Wet Weight Changes (Biological Replicates)",
         subtitle = "Averaged across technical replicates, split by growth status",
         x = "Species",
         y = "Wet Weight Change (%)",
         fill = "Growth Status") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          legend.position = "bottom")
  
  # Summary statistics by growth status
  summary_stats <- wet_weight_exp1 %>%
    group_by(growth_status) %>%
    summarise(
      n = n(),
      mean_weight_change = mean(weight_percent_change_mean, na.rm = TRUE),
      sd_weight_change = sd(weight_percent_change_mean, na.rm = TRUE),
      median_weight_change = median(weight_percent_change_mean, na.rm = TRUE)
    )
  
  print(p5)
  
  # Print summary table
  kable(summary_stats,
        caption = "**Weight Change Summary by Growth Status**",
        col.names = c("Growth Status", "N", "Mean Weight Change (%)", "SD", "Median"),
        digits = 2)
  
} else {
  plot.new()
  text(0.5, 0.5, "No wet weight data available for Experiment 1_1", cex = 1.2)
}
```

## Experiment 1_2 Weight Changes

### Wet Weight Changes

```{r wet-weight-exp2, fig.height=8}
wet_weight_exp2 <- exp_data_avg %>%
  filter(experiment == "1_2", !is.na(weight_0dpi_mean), !is.na(weight_14dpi_mean))

if(nrow(wet_weight_exp2) > 0) {
  # Create plot with growth status
  p6 <- ggplot(wet_weight_exp2, aes(x = factor(species, levels = rev(species_order)), y = weight_percent_change_mean)) +
    geom_boxplot(aes(fill = growth_status), alpha = 0.7, position = position_dodge(0.8)) +
    geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
    scale_fill_manual(values = c("Control" = "gray50", "Growth Observed" = "lightsteelblue", "No Growth" = "lightcoral")) +
    coord_flip() +
    labs(title = "Experiment 1_2: Wet Weight Changes (Biological Replicates)",
         subtitle = "Averaged across technical replicates, split by growth status",
         x = "Species",
         y = "Wet Weight Change (%)",
         fill = "Growth Status") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          legend.position = "bottom")
  
  # Summary statistics by growth status
  summary_stats <- wet_weight_exp2 %>%
    group_by(growth_status) %>%
    summarise(
      n = n(),
      mean_weight_change = mean(weight_percent_change_mean, na.rm = TRUE),
      sd_weight_change = sd(weight_percent_change_mean, na.rm = TRUE),
      median_weight_change = median(weight_percent_change_mean, na.rm = TRUE)
    )
  
  print(p6)
  
  # Print summary table
  kable(summary_stats,
        caption = "**Weight Change Summary by Growth Status**",
        col.names = c("Growth Status", "N", "Mean Weight Change (%)", "SD", "Median"),
        digits = 2)
  
} else {
  plot.new()
  text(0.5, 0.5, "No wet weight data available for Experiment 1_2", cex = 1.2)
}
```

Overall samples with fungal growth showed reduced wet weights after 14 days of incubation at 25°C 80% humidity. This weight loss is probably due to the fungal enzymes enabling faster break down and making more nutrients easily available for bacteria and the fungi.

# Summary of Replicate Structure and Growth Status Effects

```{r summary-table, results='asis'}
# Comprehensive summary of experimental design and outcomes
comprehensive_summary <- exp_data %>%
  group_by(species, experiment, growth_status) %>%
  summarise(
    biological_replicates = n_distinct(biological_replicate),
    technical_replicates = n(),
    avg_tech_per_bio = round(n() / n_distinct(biological_replicate), 1),
    .groups = "drop"
  ) %>%
  arrange(species, experiment, growth_status)

# Overall summary by growth status
growth_summary <- exp_data %>%
  group_by(growth_status) %>%
  summarise(
    total_samples = n(),
    unique_species = n_distinct(species),
    biological_replicates = n_distinct(biological_replicate),
    experiments = paste(unique(experiment), collapse = ", ")
  )

# Print tables
cat("## Experimental Design Summary\n\n")
kable(comprehensive_summary,
      caption = "**Replicate Structure by Species, Experiment, and Growth Status**",
      col.names = c("Species", "Experiment", "Growth Status", "Biological Reps", "Technical Reps", "Avg Tech/Bio"))

cat("\n\n")

kable(growth_summary,
      caption = "**Overall Summary by Growth Status**",
      col.names = c("Growth Status", "Total Samples", "Species", "Biological Reps", "Experiments"))
```

**Key Insights from Growth Status Analysis:**

-   **Growth Success**: Variability in fungal growth success across species highlights the importance of strain selection
-   **pH Dynamics**: Samples with successful fungal growth consistently show higher pH increases (\~3-4 units) compared to no-growth samples
-   **E. coli Effects**: Fungal growth generally leads to negative log reduction (increase) in E. coli concentrations, supporting the hypothesis that fungi make nutrients more accessible to bacteria\
-   **Weight Loss**: Successful fungal growth correlates with greater weight loss, indicating enhanced degradation activity
-   **Statistical Power**: Averaging technical replicates provides more robust estimates of biological variation and treatment effects

What has been achived so far

-   Identified some isolates and strains that did show reliably growth on pure faecal material and some that showed reliable growth on mixed faecal material.

-   Reached faster material degradation progressing faster into the alkaline phase of organic material degradation

-   Fungi increase bacterial concentrations and acitvity

Lessons learned

-   All tested fungi in this setting will lead to the same results at some point carbon becomes rare and growth will stop. degradation of aminoacids will produce ammonia that increases pH to levels that slow down degradation of the fungus because the enzymes denature and are not active anymore. This means also the ligno cellulose degradation will stop or slow down dramatically.

-   The 2:1 wood cips samples have a balanced out C:N ratio but the carbon is locked away in the wood and degradation is to slow to stop the increase of the pH value. Carbon needs to be easily accesible for the fungus to make it thrive, presence of carbon alone is not enough.

-   To keep the fungi and their enzyme active longer "easy" accessible carbon is needed in the bio mass. More of the proteins could then be bound in to fungal biomass instrewad of being degraded by bacteria to ammonia. This could lead to higher production of konidia ( in-case of Trichoderma), become easily degradable biomass for a further composting step or being dried and used as fertilizer.

What are the next goals

-   Trying to create a medium that stays in acidic conditions longer, to keep fungi dominant over bacteria degradation to turn more of the sample into fungal biomass and keep degradation rate high until it is significantly degraded and the pH does not increase but stays in a more neutral range.

    How will it be tested

    -   Wood chips used are resistant to degradation due to the high lignin content and the high material denisity

    -   More easily degradable carbon sources could help keeping the fungus longer active and producing acids that keep the environment acidic for longer

    -   Materials:

        -   Flax straw lower in lignin, pre-treated due to gaining of flax fibres for fabric industry

        -   Cellulose powder no lignin just simple very available form of cellulose, to dense for usage alone would not provide enough aeration

    -   Fungal strains used

        -   Sordaria one species

        -   Trichoderma koningii

The trial will contain wood chips mixed faeces, flax straw mixed faeces and flax straw and cellulose powder mixed faeces

each treatment will be run in triplicates using 15 grams of faeces in each sampled every 7 days and checked for pH and bacterial concentrations

measure procentual weight loss after each week weight before sampling and after sampling

Expected results are pH is kept lower in both flax straw and flax straw mixed with cellulose powder treatments. Cellulose powder will even stay longer acidic compared to flax straw alone.

Higher weight loss due to degradation of faecal material and faster degradation of the additives

bacterial concentrations might stay more acitve longer due to balanced pH values, acidic conditions might eliminate e coli like in untreated samples.

Goal develop an assay that the fungus can keep the pH down for more than three weeks.
